Untargeted control doc

### 2023-09-30 untargeted/paruntargeted_control.Rmd
Adding headers, isotope exploration, and multivariate analysis code

xcms_file_data <- parametadata %>%
  filter(amendment==!!amendment) %>%
  filter(polarity==!!polarity) %>%
  filter(IS_type==!!IS_type) %>%
  mutate(samp_group=paste(amendment, depth, startime, timepoint, sep="-"))
```

@@

```{r rescue feats rendering}
anno_manu <- readxl::read_excel("untargeted/manual_annotations.xlsx") %>%
  filter(polarity==!!polarity) %>%
  filter(amendment==!!amendment)
all_feat_data <- read_csv(paste0(output_folder, "all_feat_data.csv"))
anno_manu %>%
  filter(str_detect(notes, "[Rr]escue"))
all_feat_data %>%
  filter(mzmed%between%pmppm(132.1025, 10))
```

@@

manual_anno <- readxl::read_excel("untargeted/manual_annotations.xlsx")
manual_anno %>% filter(str_detect(notes, "Unpicked"))
```

@@

@@

iso_areas <- iso_mzs %>%
  split(.$filename) %>% 
  imap(function(file_peaks, file_name_i){
    msdata <- open_dataset("tmzMLs/pqds") %>%
      filter(amendment=="Amm") %>%
      filter(polarity=="pos") %>%
      filter(filename==file_name_i) %>%
      dplyr::collect()
    file_peaks %>%
      pmap(function(...){
        row_data <- data.frame(...)
        area <- msdata[mz%between%pmppm(row_data$labeled_mzmed, 100)] %>%
          .[rt%between%c(row_data$rtmin/60, row_data$rtmax/60)] %>%
          .[,sum(int)]
        cbind(row_data, area=area)
      }) %>%
      bind_rows()
  }, .progress=TRUE) %>%
  bind_rows() %>%
  select(compound_name, filename, area, n_count)

output_csv_write(iso_areas)

iso_areas %>%
  filter(compound_name=="L-Arginine") %>%
  left_join(parametadata) %>%
  filter(!is.na(depth)) %>%
  mutate(timepoint=factor(timepoint, levels=paste0("T", c(0,1,3,10,26)))) %>%
  ggplot() +
  geom_boxplot(aes(x=factor(n_count), y=area, color=timepoint)) +
  facet_grid(depth~startime, scales="free_y")
```

>>>>

xcms_file_data <- parametadata %>%
  filter(amendment==!!amendment) %>%
  filter(polarity==!!polarity) %>%
  filter(IS_type==!!IS_type) %>%
  mutate(samp_group=paste(amendment, depth, startime, timepoint, sep="-"))

mz_group <- function(mz_vals, ppm){
  group_vec <- numeric(length(mz_vals))
  group_num <- 1L
  init_vec <- mz_vals
  names(init_vec) <- seq_along(init_vec)
  while(length(init_vec)>0){
    mz_i <- init_vec[1]
    err <- mz_i*ppm/1000000
    mz_idxs <- init_vec>mz_i-err & init_vec<mz_i+err
    group_vec[as.numeric(names(mz_idxs)[mz_idxs])] <- group_num
    init_vec <- init_vec[!mz_idxs]
    group_num <- group_num+1L
  }
  group_vec
}
```

# XCMS things

@@

# Annotation

```{r rescue feats rendering, eval=FALSE}
anno_manu <- readxl::read_excel("untargeted/manual_annotations.xlsx") %>%
  filter(polarity==!!polarity) %>%
  filter(amendment==!!amendment)
all_feat_data <- read_csv(paste0(output_folder, "all_feat_data.csv"))
anno_manu %>%
  filter(str_detect(notes, "[Rr]escue"))
all_feat_data %>%
  filter(mzmed%between%pmppm(132.1025, 10))
```

@@

manual_anno <- readxl::read_excel("untargeted/manual_annotations.xlsx")
# manual_anno %>% filter(str_detect(notes, "Unpicked"))
```

@@

# Isotope exploration

@@

iso_areas <- iso_mzs %>%
  split(.$filename) %>% 
  imap(function(file_peaks, file_name_i){
    msdata <- open_dataset("tmzMLs/pqds") %>%
      filter(amendment==!!amendment) %>%
      filter(polarity==!!polarity) %>%
      filter(filename==file_name_i) %>%
      dplyr::collect()
    file_peaks %>%
      pmap(function(...){
        row_data <- data.frame(...)
        area <- msdata[mz%between%pmppm(row_data$labeled_mzmed, 10)] %>%
          .[rt%between%c(row_data$rtmin/60, row_data$rtmax/60)] %>%
          .[,sum(int)]
        cbind(row_data, area=area)
      }) %>%
      bind_rows()
  }, .progress=TRUE) %>%
  bind_rows() %>%
  mutate(labeled_n_count=n_count-1) %>%
  select(compound_name, filename, area, labeled_n_count)

output_csv_write(iso_areas)
```

```{r individual isotope EIC}
single_cmpd_iso_req <- iso_mzs %>%
  filter(compound_name=="Taurine") %>%
  distinct(compound_name, n_count, labeled_mzmed) %>%
  mutate(labeled_mzmed=round(labeled_mzmed, 5)) %>%
  summarise(req_cmd=paste0("mz%between%pmppm(", labeled_mzmed, ", 10)", collapse="|")) %>%
  pull(req_cmd)
full_arrow_req <- paste(
  'open_dataset("tmzMLs/pqds") %>%',
  'filter(amendment==!!amendment) %>%',
  'filter(samp_type=="Smp") %>%',
  'filter(polarity==!!polarity) %>% ',
  'filter(', single_cmpd_iso_req, ') %>% ',
  'dplyr::collect()'
)
single_cmpd_iso_eic <- eval(parse(text=full_arrow_req))

iso_mzs %>%
  distinct(compound_name, feature, n_count, labeled_mzmed) %>%
  filter(compound_name=="Taurine") %>%
  pmap(function(...){
    row_data <- list(...)
    single_cmpd_iso_eic %>%
      filter(mz%between%pmppm(row_data$labeled_mzmed, 10)) %>%
      mutate(n_count=row_data$n_count)
  }) %>%
  bind_rows() %>%
  # filter(rt%between%c(15, 17)) %>%
  left_join(parametadata) %>%
  filter(depth=="Surf" & startime=="Morn") %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename, color=timepoint)) +
  facet_grid(timepoint~n_count)
```

```{r individual isotope area plots}
iso_areas <- read_csv(paste0(output_folder, "iso_areas.csv"))

cmpd_name_i <- "Taurine"
single_cmpd_iso_data <- iso_areas %>%
  filter(compound_name==cmpd_name_i) %>%
  left_join(parametadata) %>%
  filter(!is.na(depth)) %>%
  mutate(labeled_n_count=factor(labeled_n_count))

single_cmpd_iso_data %>%
  ggplot() +
  geom_boxplot(aes(x=labeled_n_count, y=area, color=timepoint)) +
  facet_grid(startime~depth)

single_cmpd_iso_data %>%
  group_by(depth, startime, labeled_n_count) %>%
  mutate(area=area/sum(area)) %>%
  ggplot() +
  geom_boxplot(aes(x=labeled_n_count, y=area, color=timepoint)) +
  facet_grid(startime~depth)

single_cmpd_iso_data %>%
  group_by(depth, startime, labeled_n_count, timepoint) %>%
  summarize(area=median(area)) %>%
  group_by(depth, startime, timepoint) %>%
  mutate(lower_bound=cumsum(lag(area, default = 0))/sum(area)) %>%
  mutate(upper_bound=lead(lower_bound, default = 1)) %>%
  ggplot() +
  geom_ribbon(aes(x=timepoint, ymin=lower_bound, ymax=upper_bound, 
                  fill=labeled_n_count, group=labeled_n_count)) +
  facet_grid(startime~depth) +
  scale_fill_brewer(type = "qual", palette = 3) +
  theme_bw()
```

```{r isotopes for all targ cmpds}
pdf(paste0(output_folder, "targ_cmpd_mids.pdf"), width = 6, height = 3)
walk(unique(iso_areas$compound_name), function(cmpd_name_i){
  gp <- iso_areas %>%
    filter(compound_name==cmpd_name_i) %>%
    left_join(parametadata, by="filename") %>%
    filter(!is.na(depth)) %>%
    mutate(labeled_n_count=factor(labeled_n_count)) %>%
    group_by(depth, startime, labeled_n_count, timepoint) %>%
    summarize(area=median(area), .groups = "drop") %>%
    group_by(depth, startime, timepoint) %>%
    mutate(lower_bound=cumsum(lag(area, default = 0))/sum(area)) %>%
    mutate(upper_bound=lead(lower_bound, default = 1)) %>%
    ggplot() +
    geom_ribbon(aes(x=timepoint, ymin=lower_bound, ymax=upper_bound, 
                    fill=labeled_n_count, group=labeled_n_count)) +
    facet_grid(startime~depth) +
    scale_fill_brewer(type = "qual", palette = 3) +
    labs(x="Timepoint", fill="15N count", y="Fraction of peak area in\neach isotopologue") +
    theme_bw() +
    ggtitle(cmpd_name_i)
  plot(gp)
}, .progress = TRUE)
dev.off()
```

# Multivariate approaches

```{r overall analysis}
iso_areas <- read_csv(paste0(output_folder, "iso_areas.csv"))
areas_scaled <- iso_areas %>%
  filter(labeled_n_count>0) %>%
  left_join(parametadata) %>%
  filter(!is.na(depth)) %>%
  mutate(compound_iso=paste0(compound_name, " 15N", labeled_n_count)) %>%
  arrange(desc(area)) %>%
  select(shortname, compound_iso, area) %>%
  group_by(shortname, compound_iso) %>%
  slice(1) %>%
  ungroup() %>%
  complete(shortname, compound_iso, fill = list(area=0)) %>%
  group_by(compound_iso) %>%
  # mutate(norm_area=(area-min(area))/(max(area)-min(area))) %>%
  # mutate(norm_area=rank(area)) %>%
  mutate(norm_area=scale(area)[,1]) %>%
  ungroup() %>%
  select(shortname, compound_iso, norm_area)

areas_scaled %>%
  pivot_wider(names_from = shortname, values_from = norm_area) %>%
  column_to_rownames("compound_iso") %>%
  data.matrix() %>%
  t() %>%
  prcomp() %>%
  pluck("x") %>%
  as.data.frame() %>%
  rownames_to_column("shortname") %>%
  select(shortname, PC1:PC3) %>%
  left_join(parametadata) %>%
  arrange(timepoint) %>%
  mutate(shortname=fct_inorder(shortname)) %>%
  pivot_longer(starts_with("PC")) %>%
  filter(name%in%paste0("PC", 1:3)) %>%
  ggplot() +
  geom_col(aes(x=shortname, y=value, fill=timepoint)) +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) +
  facet_wrap(~name, ncol=1)

areas_scaled %>%
  left_join(parametadata, by="shortname") %>%
  select(compound_iso, shortname, norm_area) %>%
  group_by(compound_iso) %>%
  filter(sum(norm_area==0)<n()*0.9) %>%
  pivot_wider(names_from = "compound_iso", values_from = norm_area) %>%
  column_to_rownames("shortname") %>%
  data.matrix()

```

```{r multivar analysis by depth}
multivar_analysis <- areas_scaled %>%
  left_join(parametadata, by="shortname") %>%
  nest(data=-depth) %>%
  mutate(wide_peaks=map(data, function(metab_df){
    metab_df %>%
      select(compound_iso, shortname, norm_area) %>%
      group_by(compound_iso) %>%
      filter(sum(norm_area==0)<n()*0.9) %>%
      pivot_wider(names_from = "compound_iso", values_from = norm_area) %>%
      column_to_rownames("shortname") %>%
      data.matrix()
  })) %>%
  mutate(nmds_output=map(wide_peaks, function(metab_mat){
    # env_vec <- pull(parametadata, timepoint, shortname)[rownames(metab_mat)]
    vegan::metaMDS(metab_mat, k = 2, distance = "manhattan", trace = 0, 
                   autotransform = FALSE, noshare = FALSE, wascores = FALSE)
    # vegan::MDSrotate(env_vec)
  })) %>%
  mutate(nmds_stress=map_dbl(nmds_output, pluck, "stress")) %>%
  mutate(perm_output=map(wide_peaks, function(metab_mat){
    adon_expl <- data.frame(shortname=rownames(metab_mat)) %>%
      left_join(parametadata %>% select(shortname, timepoint, startime), by="shortname")
    vegan::adonis2(
      metab_mat~timepoint+startime, data = adon_expl, by = "margin", 
      permutations = 999, method = "manhattan")
  })) %>%
  mutate(pca_output=map(wide_peaks, prcomp)) %>%
  mutate(hclust_obj=map(wide_peaks, ~hclust(dist(.x))))

multivar_analysis %>%
  mutate(nmds_points=map(nmds_output, function(nmds_output_i){
    nmds_output_i %>% pluck("points") %>% as.data.frame()
  })) %>%
  select(depth, nmds_points) %>%
  unnest_longer(nmds_points, indices_to = "shortname") %>%
  unnest_wider(nmds_points) %>%
  left_join(parametadata %>% distinct(shortname, shortname, timepoint, startime)) %>%
  # mutate(MDS2=ifelse(str_detect(depth,"Deep"), MDS2*-1, MDS2)) %>%
  ggplot(aes(label=shortname)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, fill=timepoint), color="black", size=4, pch=21) +
  facet_wrap(~depth, ncol=1) +
  theme_bw() +
  coord_fixed()

multivar_analysis$perm_output
```
